Model Order Selection and Eigen Similarity based Framework for Detection and Identification of Network Attacks

### 2016-04-13 docs/papers/2016JNCA/2016JNCA.tex
adding papers and their assets

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmcart}
\usepackage[titletoc,title]{appendix}
%%% Load packages
\usepackage{amsthm,amsmath}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
%\RequirePackage{natbib}
%\RequirePackage{hyperref}
%\usepackage[utf8]{inputenc} %unicode support
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenc}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

\usepackage{algorithm,algorithmic}% http://ctan.org/pkg/algorithms

\usepackage{natbib}

\usepackage{supertabular}
\usepackage{fancybox}
\usepackage{acronym}
\usepackage{array}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{setspace}
\usepackage{placeins}
\usepackage{longtable}
\usepackage{mathtools}
\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
\usepackage{listings}
\usepackage{epstopdf}
\usepackage{subfigure}
\usepackage{courier}
\usepackage{amsfonts}
\usepackage{morefloats}
\usepackage{lipsum}
\newcommand*\justify{%
	\fontdimen2\font=0.4em% interword space
	\fontdimen3\font=0.2em% interword stretch
	\fontdimen4\font=0.1em% interword shrink
	\fontdimen7\font=0.1em% extra space
	%\hyphenchar\font=`\-% allowing hyphenation
}

% appendix
\usepackage{hyperref}
\usepackage{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\def\includegraphic{}
%\def%\includegraphics{}

@@

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Research}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
	addressref={address1},                   % id's of addresses, e.g. {aff1,aff2}
	corref={address1},                       % id of corresponding address, if any	
	%noteref={n1},                        	% id's of article notes, if any
	email={tpbvieira@gmail.com} 				% email address
]{\inits{TPBV}\fnm{Thiago P. B.} \snm{Vieira}}
\author[
   	addressref={address1},
   	email={???????????????}
]{\inits{DFT}\fnm{Danilo F.} \snm{Tenório}}
\author[
   	addressref={address1},
   	email={???????????????}
]{\inits{JPCLC}\fnm{João Paulo C. L.} \snm{da Costa}}
\author[
   	addressref={address1,address2},
   	email={???????????????}
]{\inits{EPF}\fnm{Edison P.} \snm{de Freitas}}
%\author[
%   	addressref={address3},
%   	email={???????????????}
%]
%{\inits{CHC}\fnm{\\Carlos H.} \snm{Cardonha}}
\author[
   	addressref={address1},
   	email={???????????????}
]{\inits{RTSJ}\fnm{\\Rafael T.} \snm{de Sousa Júnior}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=address1]{                           								% unique id
	\orgname{Department of Electrical Engineering, University of Brasilia (UnB)},	% 	university, etc
%	\street{},                     												%
	\postcode{70910-900}                                							% post or zip code
	\city{Brasília-DF},                              							% city
	\cny{Brazil}                                    								% country  
}
\address[id=address2]{%
	\orgname{Graduate Program in Electrical Engineering, Federal University of Rio Grande do Sul (UFRGS)},
%	\street{},
	\postcode{90035-190}
	\city{Porto Alegre},
	\cny{Brazil}
}
%\address[id=address3]{                           								% unique id
%	\orgname{IBM Research},	% 	university, etc
%	\street{},                     												%
%	\postcode{}                                							% post or zip code
%	\city{São Paulo},                              							% city
%	\cny{Brazil}                                    								% country  
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{artnotes}
	%\note{Sample of title note}     	% note to the article
	%\note[id=n1]{Equal contributor} 		% note, connected to author
\end{artnotes}

\end{fmbox}								% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract} % abstract

Novel schemes for attack detection are crucial to identify adaptive malicious traffic coming from sources that are quickly mobilized by attackers in high throughput communication networks. In this context, signal processing techniques have been applied to attack detection due to their capability to detect anomalies that are previously unknown, i.e. blind detection. This paper proposes a signal processing framework for the detection and identification of network attacks using concepts of model order selection (MOS), eigenvalues and similarity analysis. In order to validate our proposed framework, we consider a network traffic sample that contains malicious activity such as denial of service (DoS) and port probing. We propose to model the network traffic as a superposition of components, namely, user's operations (legitimate traffic), network service operation not related to the user (noise) and the malicious activity. The experiments performed in a real network show that the proposed blind detection approach achieves satisfactory levels of accuracy in terms of timely detection and identification of TCP/UDP ports under attack.

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
	\kwd{Network Attack Detection}
	\kwd{Model Order Selection}
	\kwd{Eigen Analysis}
	\kwd{Similarity Analysis}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section{Introduction}
\label{sec:introduction}

@@

This paper is organized as follows. In Section \ref{sec:relatedworks}, related works are discussed. Section \ref{sec:datamodel} presents the data model. Section \ref{sec:prop_getv} describes our proposed framework for blind and automatic detection of malicious traffic. Section \ref{sec:experimentalresults} discusses the experimental validation and presents the corresponding experimental results. Section \ref{sec:conclusionandfutureworks} draws the conclusions and the suggestions for future work. The Appendix \ref{sec:mos} presents mathematical concepts of examples of state-of-the-art MOS schemes.

@@

Data mining is often used to describe the process of extracting useful information from large databases. Multiple methods of data mining are used in \cite{he2008applying} to analyze data flow in a network with the aim of identifying characteristics of malicious traffic in large scale environments. Researchers have applied data mining techniques in log analysis \cite{ghourabi2010data} to improve intrusion detection performance. However, data mining techniques require prior collection of large data sets, which is a weakness of several schemes for online or low latency analysis \cite{Huang2009}.

Regular file analysis \cite{raynal2004honeypot} consists of using traffic analysis for detecting known patterns that indicate the presence of specific attacks, applying statistical analysis to the study of collected traffic. An essential feature of this method is that it depends on prior knowledge of the details of the attacks to be identified, and also depends on previous log collection for applying traffic analysis and reducing false positives.

@@

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work did not address problems without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work was not applied to port scan and denial of service (DoS) attack detection, simultaneously.

@@

Lee \emph{et al} \cite{Lee2013} proposed OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. In contrast to Lee \emph{et al}, our framework applies MOS for detection of time frames under attack and similarity analysis to extract details for detection of time and ports under attack. Additionally, Lee \emph{et al} only evaluated their proposed scheme for covariance analysis, while we adopted covariance and correlation analysis for DoS and probing attacks, respectively.

@@

In this paper the scalars are denoted by italic letters (\emph{a, b, A, B, $α$, $β$}), vectors by lowercase bold letters (\textbf{a, b}), matrices by uppercase bold letters (\textbf{A, B}), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix \textbf{A}. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. We define the operator $\rm diag(\cdot)$ that returns the vector of the main diagonal of a given matrix, the operator $\rightarrow$, which denotes the deletion of a given element from a set and the operator $\#$, that returns the rank of a matriz.

This section presents details of the simulated scenario along with a description of the dataset model as a signal superposition of legitimate traffic, noise and malicious traffic. The Subsection \ref{sec:DataCollection} describes the environment and scenario adopted in order to reproduce DoS and probing attacks. The Subsection \ref{sec:ModelingData} presents how network traffic can be modelled as signal superposition, and the Subsection \ref{sec:AnalyzedScenario} details the network traffic of synflood, fraggle and port scan attacks.

@@

%\begin{figure}[h!]
%     \centering 
%     \includegraphics[height=5cm, width=7cm]{results/figures/fig09.png}
%     \caption{\csentence{Analyzed Scenario.} Scenario to reproduce legitimate traffic, noise, DoS and port scan.}
%     \label{fig:fig1}
%\end{figure}

Note that the set of network traffic is modeled as legitimate, noise and malicious traffic, where the victim performs legitimate activities, that can be characterized by web access. In many organizations this type of traffic is predominant, since most of corporate services are web-based, such as: web pages, customized web-based systems and cloud services. It is possible to characterize the traffic of a DHCP service as an example of noise associated with the transport layer. For malicious traffic, three types of networks attacks are evaluated: synflood, fraggle and port scan. These attacks are reproduced using well-known security tools, such as Nmap to port scan, Metaspoit to synflood and Hping to lead the fraggle attack.

@@

In the proposed framework, the goal is to detect the anomalies only taking into account the traffic profile, i.e. specific information such as origin IP or day and time of the attack are not considered. Therefore, from the whole log information, we just consider the timestamp (for sequencing), port type and port number.

@@

Modeling the dataset as a signal superposition, the network traffic (\textbf{X}) can be characterized as a mixture of three components: legitimate traffic (\textbf{S}), noise (\textbf{N}) and malicious traffic (\textbf{A}), according to the following expression:

\begin{equation}\label{eq:eq01}
\boldsymbol{X}^{(q)} = \boldsymbol{U}^{(q)} + \boldsymbol{N}^{(q)} + \boldsymbol{A}^{(q)},
\end{equation}
where $q$ represents the $q$-th time frame, which is a time grouping of network traffic.

The matrix $\boldsymbol{X}^{(q)} \in \mathbb{R}^{M \times N}$ consists of \emph{M} rows and \emph{N} columns. Each row represents a communication port (TCP port or UDP port), and each column represents the time, whose unit can be minutes, for instance. Each element $x_{m,n}^{(q)}$ stands for the number of times that the port $m$ appears at the $n$-th minute, at the $q$-th time frame.

The legitimate traffic $\boldsymbol{U}^{(q)}$ is characterized by the traffic from user's operations. When a user accesses a web page, for example, there is the corresponding TCP/IP traffic to request the page, as well as there is the traffic required to domain name resolution. Figure \ref{fig:fig3} depicts an example of the legitimate traffic obtained during experiments, represented by traffic to the ports 67 and 68, which are related to the Bootstrap Protocol and to normal DHCP operations, respectively.

%\begin{figure}[h!]
%     \centering 
%     \includegraphics[height=4.5cm, width=9cm]{results/figures/fig03.png}
%     \caption{\csentence{Legitimate Traffic.} The traffic from user's operations, that can be characterized by web access, traffic of well-known applications or network protocols.}
%     \label{fig:fig3}
%\end{figure}

All traffic, that is not directly associated with users' operations, but it is not a malicious traffic, is considered as noise $\boldsymbol{N}^{(q)}$. The automatic acquisition service of logical IP network address (DHCP) is an example of noise. Independently of any user operation, the machine receives an IP address, since it is configured to perform a DHCP address request. Figure \ref{fig:fig4} depicts an example of noise in a network traffic.

%\begin{figure}[h!]
%     \centering 
%     \includegraphics[height=4.5cm, width=9cm]{results/figures/fig04.png}
%     \caption{\csentence{Noise Traffic.} Network traffic of user independent operations for network management.}
%     \label{fig:fig4}
%\end{figure}

The traffic coming from a malicious activity, such as a synflood or fraggle attack, is represented by the matrix $\boldsymbol{A}^{(q)}$. For this work we only consider the traffic from port scanning and flood attacks, which aims to cause denial of service.

@@

The TCP protocol is a connection-oriented protocol, then a virtual connection must be established between two computers for an end-to-end TCP communication. This virtual connection requires a ``handshake'', that occurs in three steps, known as three-way handshake. If a computer needs to communicate with another computer, the requester sends a packet communication synchronization (SYN) to a specific destination port, which is in listening state. If the destination is active, running and accepting requests, it responds to the requester with a SYN/ACK confirmation message. After receiving this message, the requester sends an ACK message to the destination and then the connection is established.

@@

%\begin{figure}[h!]
%     \centering 
%     \includegraphics[height=4.5cm, width=9cm]{results/figures/fig05.png}
%     \caption{\csentence{Synflood Traffic.} A large quantity of SYN requests to a target, in order to cause a DoS.}
%     \label{fig:fig5}
%\end{figure}

With respect to the fraggle atack, large packets with ``UDP echo'' segments are sent to the broadcast address of a network. Every packet is modified to have the source address of the victim, in order to implement the source address spoofing technique. Therefore, each host receives a huge amount of requests ``UDP echo'' and all of them replies to the IP address of the victim, causing a packet flooding aiming a DoS. This attack can affect the entire network, since all hosts receive several requests ``UDP echo'' and respond with the ICMP protocol, therefore each host acts as an ``amplifier'' of the attack. This last part of the fraggle attack is not taken into account in this work, because the victim receives ICMP (network layer) packets originated from the hosts that are attacked with flooding packet ``UDP echo''. This occurs due to the UDP be not able to know if the segment sent has reached its destination.

Figure \ref{fig:fig6} depicts an example of the fraggle attack in a real computer network. More than 6,000,000 malicious packets can be counted in an interval of ten minutes, which can be considered an abnormal network traffic, especially due to the concentrated traffic in a short period of time and due to the similarity of the outstanding traffic.

%\begin{figure}[h!]
%     \centering 
%     \includegraphics[height=4.5cm, width=9cm]{results/figures/fig06.png}
%     \caption{\csentence{Fraggle Traffic.} Large amount of “UDP echo” requests and replies, causing packet flooding.}
%     \label{fig:fig6}
%\end{figure}

@@

In TCP SYN scan, a SYN packet is sent to the destination and two types of responses may occur: SYN/ACK or RST/ACK. In the first case, the destination port is in the listening state, in the second case, the destination port is not listening. At the end of each port scanning, a RST/ACK packet is sent by the system that is performing the port  scan. Therefore, a full connection or a complete three-way handshake is never established, which makes the detection of the attack sender more difficult, and requires approaches able to identify probing attacks without connection stablishment. The UDP scan technique sends UDP packets to the destination port, and if it responds with a ``ICMP port unreachable'' message, then it indicates that the scanned port is closed. On the other hand, if a message is not received, then the port is considered as open.

%\begin{figure}[h!]
%     \centering 
%     \includegraphics[height=5cm, width=9cm]{results/figures/fig07.png}
%     \caption{\csentence{Port Scan Traffic.} Connection attempts in order to identify active ports.}
%     \label{fig:fig7}
%\end{figure}

@@

The set of obtained vectors $\boldsymbol{y}_{m}^{(q)}$ composes the zero mean matrix $\boldsymbol{Y}^{(q)}$, then the covariance matrix $\boldsymbol{R}_{yy}^{(q)}$ can be calculated as follows

>>>>

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{algorithm,algorithmic}% http://ctan.org/pkg/algorithms
\usepackage{natbib}
\usepackage{supertabular}
\usepackage{fancybox}
\usepackage{acronym}
\usepackage{array}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{setspace}
\usepackage{placeins}
\usepackage{longtable}
\usepackage{mathtools}
\usepackage{listings}
\usepackage{epstopdf}
\usepackage{subfigure}
\usepackage{courier}
\usepackage{amsfonts}
\usepackage{morefloats}
\usepackage{lipsum}
\usepackage{amsthm,amsmath}
\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand*\justify{%
	\fontdimen2\font=0.4em% interword space
	\fontdimen3\font=0.2em% interword stretch
	\fontdimen4\font=0.1em% interword shrink
	\fontdimen7\font=0.1em% extra space
	%\hyphenchar\font=`\-% allowing hyphenation
}

@@

% appendix
\usepackage[titletoc,title]{appendix}
\usepackage{hyperref}
\usepackage{cleveref}

\modulolinenumbers[5]

\journal{ Journal of Network and Computer Applications}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}

@@

%% Group authors per affiliation:
\author[unbaddress]{Thiago P. B. Vieira}
\author[unbaddress]{Danilo F. Tenório}
\author[unbaddress,Ilmenauaddress,Fraunhoferaddress]{João Paulo C. L. da Costa}
\author[ufrgsaddress]{\\Edison P. de Freitas}
\author[unbaddress]{Rafael T. de Sousa Júnior}

\address[unbaddress]{Department of Electrical Engineering, University of Brasilia (UnB), 70910-900, Brasília-DF, Brazil}
\address[Ilmenauaddress]{Institute for Information Technology, Ilmenau University of Technology, Ilmenau, Germany}
\address[Fraunhoferaddress]{Fraunhofer Institute for Integrated Circuits IIS, Erlangen, Germany}
\address[ufrgsaddress]{Graduate Program in Electrical Engineering, Federal University of Rio Grande do Sul (UFRGS), 90035-190, Porto Alegre, Brazil}

\begin{abstract}
Novel schemes for attack detection are crucial to identify adaptive malicious traffic coming from sources that are quickly mobilized by attackers in high throughput communication networks. In this context, signal processing techniques have been applied to attack detection due to their capability to detect anomalies that are previously unknown, i.e. blind detection. This paper proposes a signal processing framework for the detection and identification of network attacks using concepts of model order selection (MOS), eigenvalues and similarity analysis. In order to validate our proposed framework, we consider a network traffic sample that contains malicious activity such as denial of service (DoS) and port probing. We propose to model the network traffic as a superposition of components, namely, user's operations (legitimate traffic), network service operation not related to the user (noise) and the malicious activity. The experiments performed in a real network show that the proposed blind detection approach achieves satisfactory levels of accuracy in terms of timely detection and identification of TCP/UDP ports under attack.
\end{abstract}

\begin{keyword}
Network Attack Detection \sep Model Order Selection \sep Eigen Analysis \sep Similarity Analysis
\end{keyword}

@@

\linenumbers

\section{Introduction}
\label{sec:introduction}

@@

This paper is organized as follows. In Section \ref{sec:relatedworks}, related works are discussed. Section \ref{sec:datamodel} presents the data model. Section \ref{sec:prop_getv} describes our proposed framework for blind and automatic detection of malicious traffic. Section \ref{sec:experimentalresults} discusses the experimental validation and presents the experimental results. Section \ref{sec:conclusionandfutureworks} draws the conclusions and the suggestions for future work. The \ref{sec:mos} presents mathematical concepts of examples of state-of-the-art MOS schemes.

@@

Data mining is often used to describe the process of extracting useful information from large databases. Multiple methods of data mining are used in \cite{he2008applying} to analyze data flow in a network with the aim of identifying characteristics of malicious traffic in large scale environments. Researchers have applied data mining techniques in log analysis \cite{ghourabi2010data} to improve intrusion detection performance. However, data mining techniques used so far in network analysis require prior collection of large data sets, which is a limitation of several schemes for online analysis \cite{Huang2009}.

Regular file analysis \cite{raynal2004honeypot} consists of traffic analysis for detecting known patterns that indicate the presence of attacks, applying statistical analysis to the study of collected traffic. An essential feature of this method is that it depends on prior knowledge of the details of the attacks to be identified, and also depends on previous log collection for traffic analysis and false positives reduction.

@@

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work did not address problems without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work was not applied to probing and denial of service (DoS) attack detection, simultaneously.

@@

Lee \emph{et al} \cite{Lee2013} proposed OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. In contrast to Lee \emph{et al}, our framework applies MOS for detection of time frames under attack and similarity analysis to extract details for detection of time and ports under attack. Additionally, Lee \emph{et al} only evaluated their proposed scheme for covariance analysis, while we adopted an analysis based on sample covariance of zero mean variables and sample covariance of zero mean and unitary standard deviation variables, for DoS and probing attacks, respectively.

@@

In this paper, scalars are denoted by italic letters (\emph{a, b, A, B, $α$, $β$}), vectors by lowercase bold letters (\textbf{a, b}), matrices by uppercase bold letters (\textbf{A, B}), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix \textbf{A}. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. We define the operator $\rm diag(\cdot)$ that returns the vector of the main diagonal of a given matrix, the operator $\rightarrow$, which denotes the deletion of a given element from a set and the operator $\#$, that returns the rank of a matriz.

This section presents details of the simulated scenario along with a description of the dataset model as a signal superposition of legitimate traffic, noise and malicious traffic. Subsection \ref{sec:DataCollection} describes the environment and scenario adopted in order to reproduce DoS and probing attacks. Subsection \ref{sec:ModelingData} presents how network traffic can be modelled as signal superposition, and Subsection \ref{sec:AnalyzedScenario} details the traffic of synflood, fraggle and port scan attacks.

@@

\begin{figure}[h!]
     \centering 
     \includegraphics[height=5cm, width=7cm]{results/figures/fig09.png}
     \caption{Scenario to reproduce legitimate traffic, noise, DoS and port scan.}
     \label{fig:fig1}
\end{figure}

Note that the set of network traffic is modeled as legitimate, noise and malicious traffic, where the victim performs legitimate activities, that can be characterized by web access. In many organizations this type of traffic is predominant, since most of corporate services are web-based, such as: web pages, customized web-based systems and cloud services. It is possible to characterize the traffic of a DHCP service as an example of noise associated with the transport layer. For malicious traffic, three types of networks attacks are evaluated: synflood, fraggle and port scan. These attacks are reproduced using well-known security tools, such as Nmap\footnote{http://nmap.org} to port scan, Metasploit\footnote{http://www.metasploit.com} to synflood and Hping\footnote{http://hping.org} to lead the fraggle attack.

@@

In the proposed framework, the goal is to detect the anomalies only taking into account the traffic profile, i.e. specific information such as origin IP or day and time of the attack are not considered. Therefore, from the entire log information, we just consider the timestamp (for sequencing), port type and port number.

@@

By modeling the dataset as a signal superposition, the network traffic (\textbf{X}) can be characterized as a mixture of three components: legitimate traffic (\textbf{S}), noise (\textbf{N}) and malicious traffic (\textbf{A}), according to the following expression:

\begin{equation}\label{eq:eq01}
\boldsymbol{X}^{(q)} = \boldsymbol{U}^{(q)} + \boldsymbol{N}^{(q)} + \boldsymbol{A}^{(q)},
\end{equation}
where $q$ represents the $q$-th time frame, which is a time grouping of network traffic. The matrix $\boldsymbol{X}^{(q)} \in \mathbb{R}^{M \times N}$ consists of \emph{M} rows and \emph{N} columns. Each row represents a communication port (TCP port or UDP port), and each column represents the time, whose unit can be minutes, for instance. Each element $x_{m,n}^{(q)}$ stands for the number of times that the port $m$ appears at the $n$-th minute, at the $q$-th time frame.

The legitimate traffic $\boldsymbol{U}^{(q)}$ is characterized by the traffic from user's operations. When a user accesses a web page, for example, there is the corresponding TCP/IP traffic to request the page, as well as there is the traffic required to domain name resolution. Figure \ref{fig:fig3} depicts an example of the legitimate traffic obtained during experiments.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=4.5cm, width=9cm]{results/figures/fig03.png}
     \caption{Traffic from user's operations, that can be characterized by web access, traffic of well-known applications or network protocols.}
     \label{fig:fig3}
\end{figure}

The traffic that is not associated with user's operations and with malicious traffic is modelled as noise $\boldsymbol{N}^{(q)}$. The automatic acquisition service of logical IP network address (DHCP) is an example of noise. Independently of any user operation, the machine receives an IP address, since it is configured to perform a DHCP address request. Figure \ref{fig:fig4} depicts an example of noise in a network traffic, represented by traffic to ports 67 and 68.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=4.5cm, width=9cm]{results/figures/fig04.png}
     \caption{Network traffic of user independent operations for network management.}
     \label{fig:fig4}
\end{figure}

The traffic coming from a malicious activity, such as a synflood or fraggle attack, is represented by the matrix $\boldsymbol{A}^{(q)}$. For this work we only consider the traffic from port scanning and flood attacks.

@@

@@

\begin{figure}[h!]
     \centering 
     \includegraphics[height=4.5cm, width=9cm]{results/figures/fig05.png}
     \caption{A large quantity of SYN requests to a target, in order to cause a DoS.}
     \label{fig:fig5}
\end{figure}

With respect to the fraggle atack, large packets with UDP echo segments are sent to the broadcast address of a network. Every packet is modified to have the source address of the victim, in order to implement the source address spoofing technique. Therefore, each host receives a huge amount of requests UDP echo and all of them replies to the IP address of the victim, causing a packet flooding aiming a DoS. This attack can affect the entire network, since all hosts receive several requests UDP echo and respond with the ICMP protocol, therefore each host acts as an amplifier of the attack. This last part of the fraggle attack is not taken into account in this work, because the victim receives ICMP (network layer) packets originated from the hosts that are attacked with flooding packet UDP echo. Figure \ref{fig:fig6} depicts an example of the fraggle attack in a real computer network.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=4.5cm, width=9cm]{results/figures/fig06.png}
     \caption{Large amount of “UDP echo” requests and replies, causing packet flooding.}
     \label{fig:fig6}
\end{figure}

More than 6,000,000 malicious packets can be counted in an interval of ten minutes, which can be considered an abnormal network traffic, especially due to the concentrated traffic in a short period of time and due to the similarity of the outstanding traffic.

@@

In TCP SYN scan, a SYN packet is sent to the destination and two types of responses may occur: SYN/ACK or RST/ACK. In the first case, the destination port is in the listening state, in the second case, the destination port is not listening. At the end of each port scanning, a RST/ACK packet is sent by the system that is performing the port  scan. Therefore, a full connection or a complete three-way handshake is never established, which makes the detection of the attack sender more difficult, and requires approaches able to identify probing attacks without connection stablishment. The UDP scan technique sends UDP packets to the destination port, and if it responds with a \emph{ICMP port unreachable} message, then it indicates that the scanned port is closed. On the other hand, if a message is not received, then the port is considered as open.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=5cm, width=9cm]{results/figures/fig07.png}
     \caption{Connection attempts in order to identify active ports.}
     \label{fig:fig7}
\end{figure}

@@

The set of obtained vectors $\boldsymbol{y}_{m}^{(q)}$ composes the zero mean matrix $\boldsymbol{Y}^{(q)}$, then the sample covariance matrix $\boldsymbol{\hat{R}}_{yy}^{(q)}$ can be calculated as follows
